Modelling BCG vaccination in the UK

### 2019-07-04 rmds/04-tb-epi-england.Rmd
checked accroynm usage

Whilst the characteristics of tuberculosis in England have been reported elsewhere,[@PHE2017; @PHE2018] and key risk factors such as non-UK birth status have been identified,@French2007 little attention has been given to the role of BCG vaccination. In particular, there is little information available regarding the demographics of vaccinated versus unvaccinated cases and the impact of BCG vaccination on TB outcomes in England has not been explored. There has also only been limited reporting of the age distribution, and trends over time, in incidence rates stratified by UK birth status.

@@

The enhanced tuberculosis surveillance system (ETS) is maintained by Public Health England (PHE) and collects demographic, clinical, and microbiological data on all notified cases in England. Notification is required by law, with health service providers having to inform PHE of all confirmed TB cases.[@PHE2017] Data collection began in 2000 and was expanded, with additional variables, with the launch of a web based system in 2008.[@Kriujshaar2007] It is updated annually with de-notifications, late notifications and other updates. A descriptive analysis of tuberculosis epidemiology in England is published each year, which reports on data collection, cleaning, and trends in TB incidence at both a national, and sub-national level.[@PHE2017]

@@

Data on all notifications from the Enhanced Tuberculosis Surveillance (ETS) system from 2000 to 2015 were obtained from PHE. Data fields included: notification date, age, Public Health England centre, occupation, ethnic group, UK birth status, years since entry to the UK, date of symptom onset, date of presentation, date of diagnosis, date of treatment start, date of treatment end, date of death, Pulmonary TB status, culture status, sputum smear status, drug resistance, BCG vaccination status, year of vaccination, outcome at 12 months, overall outcome, and cause of death. Invalid entries were replaced with missing values unless otherwise noted, with character variables stored as factors using their most common entry as the baseline. Notifications from Scotland, Northern Ireland and Wales were dropped from the data-set. Several variables were created, or modified, for use in further analysis, Table \@ref(tab:created-var) summarises these variables. The code used for data cleaning is available as an R package ([https://www.samabbott.co.uk/tbinenglanddataclean/reference/clean_munge_ets_2016.html](https://www.samabbott.co.uk/tbinenglanddataclean/reference/clean_munge_ets_2016.html)).

@@

pander::pander(created_var, caption = '(\\#tab:created-var) Variables derived or modified from the Enhanced Tuberculosis Surveillance system for use in the analyses throughout this thesis.', booktabs = TRUE, justify = 'left', split.cells = 50)
```

@@

The ETS is in a wide format with each notification having a single row, and with each unique variable having a single column. This structure means that the progression of tuberculosis in each individual is captured by a series of dates rather than as a series of events. As notifications are not linked to a unique patient I.D it is possible that individuals are duplicated within the ETS, with multiple notifications. These recurrent notifications have been flagged within the data extract by the TB section at PHE. The majority of variables are factors, with a significant minority of numeric and date variables.

```{r data-struct, fig.cap = "Summary plot of the structure of the Enhanced Tuberculosis Surveillance data extract used in this thesis. Due to the large size of the dataset, the data has been sub sampled with only 20\\% of the data shown. Notifications have been ordered by date of notification from left to right. Purple indicates numeric variables, green indicates factor variables, dates are shown using red and samples with missing values are shown using grey. The figure shows that the majority of the included variables are factors, with a significant minority of numeric and date variables. Data completeness is near random for some variables but has changed other time for others, for example national socioeconomic quintile (natquintile). This figure does not account for nested variables", out.width = "60%", eval = FALSE}
if (regen_results) {
  plot_data_struct <- ets %>% 
    sample_frac(0.2) %>% 
    arrange(desc(caserepdate)) %>% 
  vis_dat(warn_large_data = FALSE) + 
    coord_flip() +
    theme(legend.position = "bottom")

@@

```{r plot-missing-struct, fig.cap = "Summary plot of missing data in the extract of the Enhanced Tuberculosis Surveillance data used in this thesis. Due to the large size of the dataset, the data has been sub-sampled with only 20\\% of the data shown in this figure. Notifications have been ordered by date of notification from left to right. The following subset of variables are shown; year (year), sex (sex), age (age), Public Health England Centre (phec), Occupation (occat), Ethnic group (ethgrp), UK birth status (ukborn), Time since entry (timesinceent), date of symptom onset (symptonset), date of diagnosis (datediag), started treatment (startedtreat), date of starting treatment (starttreatdate), treatment end date (txenddate), pulmonary or extra-pulmonary TB (pulmextrapulm), culture (culture), sputum smear status (sputsmear), drug resistance (anyres), previous diagnosis (prevdiag), BCG status(bcgvacc), Year of BCG vaccination (bcgvaccyr), overall outcome (overalloutcome), cause of death (tomdeathrelate), socio-economic status quintiles (natquintile), and date of death (dateofdeath). Nested variables have been accounted for (i.e data of death has had an entry added for cases that are known to have not died), so that true missingness for all variables is estimated.", out.width = "60%"}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-missing-struct.png")
```

@@

kable(miss_ets,
      caption = "Breakdown of missing data from the ETS prior to the web based system (pre 2009) and post (post 2008) by variable, ordered by the  percentage missing for a subset of variables. The following subset of variables are shown; year (year), sex (sex), age (age), Public Health England Centre (phec), Occupation (occat), Ethnic group (ethgrp), UK birth status (ukborn), Time since entry (timesinceent), date of symptom onset (symptonset), date of diagnosis (datediag), started treatment (startedtreat), date of starting treatment (starttreatdate), treatment end date (txenddate), pulmonary or extra-pulmonary TB (pulmextrapulm), culture (culture), sputum smear status (sputsmear), drug resistance (anyres), previous diagnosis (prevdiag), BCG status(bcgvacc), Year of BCG vaccination (bcgvaccyr), overall outcome (overalloutcome), cause of death (tomdeathrelate), socio-economic status quintiles (natquintile), and date of death (dateofdeath). Nested variables have been accounted for (i.e data of death has had an entry added for cases that are known to have not died), so that true missingness for all variables is estimated.",
      booktabs = TRUE) %>%
  add_header_above(c(" ", "Pre 2009" = 2, "Post 2008" = 2)) %>% 
  kable_styling(latex_options = "HOLD_position") 
```

@@

Routine observational data-sets are subject to numerous potential biases, such as selection bias, recall bias, measurement bias, and unmeasured confounding.[@Benchimol2016a] Additionally as the data has not been collected with a specific analysis in mind there maybe issues with the specificity of variables. The ETS system is likely to suffer from all of the above biases to some extent, which must be accounted for as far as possbile, and explicitly stated at every level of analysis. The most important consideration is that the ETS system is unlikely to be representative of the general population as it contains only notified TB cases that occurred in England during the study period, research questions must therefore be either limited to active tuberculosis patients, or when extended to the general population the differing population demographics must be accounted for. If this is not done then any results may be due to selection bias. Additionally, multiple variables may suffer from misclassification bias, including BCG status which can be assessed via vaccination record, the presence of a scar, or case recall: this may lead to spurious associations.[@Fewell2007] Validation studies would be required to account for this, which is beyond the scope of this thesis.

@@

Demographic data used in this thesis is drawn from two main sources; mid-year resident populations, by single year of age, downloaded from the Office for National Statistics (ONS) website for 2000 to 2015 and population estimates from the yearly April to June Labour Force Survey (LFS) stratified by single year of age and UK birth status. The LFS is a study of the employment circumstances of the UK population and provides the official measures of employment and unemployment in the UK. It also records other details such as ethnicity and UK birth status which may be used, along with population weightings, to estimate the UK and non-UK born population.

@@

The mid-year population estimates were transformed from wide format into tidy data,[@Wickham2015] with the population estimates from 2000 being reformatted to match those from 2001 on-wards. Data from the Labour force survey was available by year, so each dataset was separately imported into R.[@R] Reporting practices have changed with time so the appropriate variables for age, country of origin, country of birth, and survey weight were extracted from each yearly extract, standardised, and combined into a single tidy dataset. The LFS data was then aggregated, accounting for survey weight, by year, age, and UK birth status to provide yearly estimates of the UK born/Non-UK born demographics by age. Finally 5 year age groups were defined using the single year of age.

@@

Both the mid-year ONS population estimates,[@storey2016] and the LFS are assessed for performance and quality elsewhere.[@NHSTraffordCCGGoverningBody2015; @Date2012] However both have several failing that it is important to note, as they could introduce bias in future analysis. Whilst the ONS mid-year, and LFS estimates compare well when aggregated by age (Figure \@ref(fig:plot-overall-pop)) there is more disagreement when they are broken down by 5 year age groups (Figure \@ref(fig:plot-age-strat)). For those at working age both data sources are comparable (with approximately a 1% difference across all years), however for children, young adults, and those who are 85+ the LFS underestimates the total population. This is particularly the case for older adults with between a 5% and 20% discrepancy for those aged 85-89 and a 25% to 45% discrepancy between those aged 90+. This could be problematic as these age groups often have the most severe outcomes to tuberculosis infection. A pragmatic approach to this is to exclude those aged 90+ from future analysis as results for this age group will be subject to large amounts of uncertainty which will be difficult to directly incorporate into the results.

```{r plot-overall-pop, fig.cap = "Overall population estimates derived using Office for National Statistics (ONS) and Labour Force Survey (LFS) demographic data. The ONS data is likely to be more reliable as the LFS data is derived using a weighted survey. After accounting for missing UK birth status both datasets provide comparable estimates of the population of England, with a clearly increasing trend over time. However the ONS data indicates a reduction in population from 2000 until 2001 that is not seen in the LFS data. The UK born population has also risen slowly from 2000 until 2015, although the biggest increase has been in the non-UK born population."}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-overall-pop.png")
```

```{r plot-age-strat, fig.cap = "Percentage difference between Office for National Statistics (ONS) population estimates and estimates derived from the Labour Force Survey (LFS) by 5 year age group. For most age groups there is less than a 2\\% difference over time. In older adults (85+) there is a substantially greater difference ranging from 5\\% to 40\\%."}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-age-strat.png")
```

## Tuberculosis notifications

@@

```{r plot-overall-nots, fig.cap = "Notifications in England from 2000 to 2015 stratified by UK birth status, sourced from the Enhanced Tuberculosis Surveillance (ETS) system. Notifications in the non-UK born doubled from 3329 in 2000 to 6021 in 2011 since when they have decreased year on year. In the UK born notifications have remained comparable over time, with some evidence of a decrease from 2011 until 2015. UK birth status has become increasingly complete over time with notifications without birth status dropping from 885 in 2000 to 121 in 2015."}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-overall-nots.png")
```

@@

```{r age-dist-cases-nots, fig.cap = "Proportion of cases by 5 year age group in the Enhanced Tuberculosis surveillance system in 2005, 2010 and 2015 stratified by UK birth status. Non-UK born cases have a higher proportion of young adult cases with very few cases in children or in older adults. UK born cases have a more uniform distribution of cases with some evidence of a higher proportion of cases in young adults. In the non-UK born the proportion of cases in young adults has decreased over time, with no evidence of a temporal trend in the UK born. These results are not adjusted for population demographics and therefore may be biased."}
if (regen_results) {

@@

## Tuberculosis incidence rates {#tb-inc-rates-epi}

@@

In the the Enhanced Tuberculosis Surveillance (ETS) system, I found a high degree of missing data for several important variables. I also found that there is likely to be strong missing at random (MAR) mechanism underlying this missing data for multiple variables. Several factors are strongly associated with data being missing for many variables, including UK birth status, ethnic group, socio-economic status and year. These MAR mechanisms must be adjusted for in future analysis to avoid bias. I found that date variables in particular suffered from changing data completeness over time, which may introduce spurious temporal trends if not fully understood. I also found that for several variables, including the date of symptom onset, that there was a large degree of recall bias both on the scales of days and months. Several variables, including date of notification and date of starting treatment showed a seasonal trend, with a maximum in the summer months. The date of ending treatment showed less evidence of a seasonal trend but there was some evidence of a maximum number of cases completing treatment in the winter months.

@@

- Tuberculosis incidence rates stratified by age and UK birth status have been calculated, along with case rates for TB outcomes. These estimates were then to extensively explore trends in TB in England, identifying possible analysis questions to be addressed later in this thesis.

>>>>

Whilst the characteristics of tuberculosis (TB) in England have been reported elsewhere,[@PHE2017; @PHE2018] and key risk factors such as non-UK birth status have been identified,@French2007 little attention has been given to the role of BCG vaccination. In particular, there is little information available regarding the demographics of vaccinated versus unvaccinated cases and the impact of BCG vaccination on TB outcomes in England has not been explored. There has also only been limited reporting of the age distribution, and trends over time, in incidence rates stratified by UK birth status.

@@

The enhanced tuberculosis surveillance system (ETS) is maintained by Public Health England (PHE) and collects demographic, clinical, and microbiological data on all notified cases in England. Notification is required by law, with health service providers having to inform PHE of all confirmed TB cases.[@PHE2017] Data collection began in 2000 and was expanded, with additional variables, with the launch of a web based system in 2008.[@Kriujshaar2007] It is updated annually with de-notifications, late notifications and other updates. A descriptive analysis of TB epidemiology in England is published each year, which reports on data collection, cleaning, and trends in TB incidence at both a national, and sub-national level.[@PHE2017]

@@

Data on all notifications from the ETS system from 2000 to 2015 were obtained from PHE. Data fields included: notification date, age, PHE centre, occupation, ethnic group, UK birth status, years since entry to the UK, date of symptom onset, date of presentation, date of diagnosis, date of treatment start, date of treatment end, date of death, Pulmonary TB status, culture status, sputum smear status, drug resistance, BCG vaccination status, year of vaccination, outcome at 12 months, overall outcome, and cause of death. Invalid entries were replaced with missing values unless otherwise noted, with character variables stored as factors using their most common entry as the baseline. Notifications from Scotland, Northern Ireland and Wales were dropped from the data-set. Several variables were created, or modified, for use in further analysis, Table \@ref(tab:created-var) summarises these variables. The code used for data cleaning is available as an R package ([https://www.samabbott.co.uk/tbinenglanddataclean/reference/clean_munge_ets_2016.html](https://www.samabbott.co.uk/tbinenglanddataclean/reference/clean_munge_ets_2016.html)).

@@

pander::pander(created_var, caption = '(\\#tab:created-var) Variables derived or modified from the ETS system for use in the analyses throughout this thesis.', booktabs = TRUE, justify = 'left', split.cells = 50)
```

@@

The ETS is in a wide format with each notification having a single row, and with each unique variable having a single column. This structure means that the progression of TB in each individual is captured by a series of dates rather than as a series of events. As notifications are not linked to a unique patient I.D it is possible that individuals are duplicated within the ETS, with multiple notifications. These recurrent notifications have been flagged within the data extract by the TB section at PHE. The majority of variables are factors, with a significant minority of numeric and date variables.

```{r data-struct, fig.cap = "Summary plot of the structure of the ETS data extract used in this thesis. Due to the large size of the dataset, the data has been sub sampled with only 20\\% of the data shown. Notifications have been ordered by date of notification from left to right. Purple indicates numeric variables, green indicates factor variables, dates are shown using red and samples with missing values are shown using grey. The figure shows that the majority of the included variables are factors, with a significant minority of numeric and date variables. Data completeness is near random for some variables but has changed other time for others, for example national socioeconomic quintile (natquintile). This figure does not account for nested variables", out.width = "60%", eval = FALSE}
if (regen_results) {
  plot_data_struct <- ets %>% 
    sample_frac(0.2) %>% 
    arrange(desc(caserepdate)) %>% 
  vis_dat(warn_large_data = FALSE) + 
    coord_flip() +
    theme(legend.position = "bottom")

@@

```{r plot-missing-struct, fig.cap = "Summary plot of missing data in the extract of the ETS data used in this thesis. Due to the large size of the dataset, the data has been sub-sampled with only 20\\% of the data shown in this figure. Notifications have been ordered by date of notification from left to right. The following subset of variables are shown; year (year), sex (sex), age (age), PHE Centre (phec), Occupation (occat), Ethnic group (ethgrp), UK birth status (ukborn), Time since entry (timesinceent), date of symptom onset (symptonset), date of diagnosis (datediag), started treatment (startedtreat), date of starting treatment (starttreatdate), treatment end date (txenddate), pulmonary or extra-pulmonary TB (pulmextrapulm), culture (culture), sputum smear status (sputsmear), drug resistance (anyres), previous diagnosis (prevdiag), BCG status(bcgvacc), Year of BCG vaccination (bcgvaccyr), overall outcome (overalloutcome), cause of death (tomdeathrelate), socio-economic status quintiles (natquintile), and date of death (dateofdeath). Nested variables have been accounted for (i.e data of death has had an entry added for cases that are known to have not died), so that true missingness for all variables is estimated.", out.width = "60%"}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-missing-struct.png")
```

@@

kable(miss_ets,
      caption = "Breakdown of missing data from the ETS prior to the web based system (pre 2009) and post (post 2008) by variable, ordered by the  percentage missing for a subset of variables. The following subset of variables are shown; year (year), sex (sex), age (age), PHE Centre (phec), Occupation (occat), Ethnic group (ethgrp), UK birth status (ukborn), Time since entry (timesinceent), date of symptom onset (symptonset), date of diagnosis (datediag), started treatment (startedtreat), date of starting treatment (starttreatdate), treatment end date (txenddate), pulmonary or extra-pulmonary TB (pulmextrapulm), culture (culture), sputum smear status (sputsmear), drug resistance (anyres), previous diagnosis (prevdiag), BCG status(bcgvacc), Year of BCG vaccination (bcgvaccyr), overall outcome (overalloutcome), cause of death (tomdeathrelate), socio-economic status quintiles (natquintile), and date of death (dateofdeath). Nested variables have been accounted for (i.e data of death has had an entry added for cases that are known to have not died), so that true missingness for all variables is estimated.",
      booktabs = TRUE) %>%
  add_header_above(c(" ", "Pre 2009" = 2, "Post 2008" = 2)) %>% 
  kable_styling(latex_options = "HOLD_position") 
```

@@

Routine observational data-sets are subject to numerous potential biases, such as selection bias, recall bias, measurement bias, and unmeasured confounding.[@Benchimol2016a] Additionally as the data has not been collected with a specific analysis in mind there maybe issues with the specificity of variables. The ETS system is likely to suffer from all of the above biases to some extent, which must be accounted for as far as possbile, and explicitly stated at every level of analysis. The most important consideration is that the ETS system is unlikely to be representative of the general population as it contains only notified TB cases that occurred in England during the study period, research questions must therefore be either limited to active TB patients, or when extended to the general population the differing population demographics must be accounted for. If this is not done then any results may be due to selection bias. Additionally, multiple variables may suffer from misclassification bias, including BCG status which can be assessed via vaccination record, the presence of a scar, or case recall: this may lead to spurious associations.[@Fewell2007] Validation studies would be required to account for this, which is beyond the scope of this thesis.

@@

Demographic data used in this thesis is drawn from two main sources; mid-year resident populations, by single year of age, downloaded from the Office for National Statistics (ONS) website for 2000 to 2015 and population estimates from the yearly April to June LFS stratified by single year of age and UK birth status. The LFS is a study of the employment circumstances of the UK population and provides the official measures of employment and unemployment in the UK. It also records other details such as ethnicity and UK birth status which may be used, along with population weightings, to estimate the UK and non-UK born population.

@@

The mid-year population estimates were transformed from wide format into tidy data,[@Wickham2015] with the population estimates from 2000 being reformatted to match those from 2001 on-wards. Data from the LFS was available by year, so each dataset was separately imported into R.[@R] Reporting practices have changed with time so the appropriate variables for age, country of origin, country of birth, and survey weight were extracted from each yearly extract, standardised, and combined into a single tidy dataset. The LFS data was then aggregated, accounting for survey weight, by year, age, and UK birth status to provide yearly estimates of the UK born/Non-UK born demographics by age. Finally 5 year age groups were defined using the single year of age.

@@

Both the mid-year ONS population estimates,[@storey2016] and the LFS are assessed for performance and quality elsewhere.[@NHSTraffordCCGGoverningBody2015; @Date2012] However both have several failing that it is important to note, as they could introduce bias in future analysis. Whilst the ONS mid-year, and LFS estimates compare well when aggregated by age (Figure \@ref(fig:plot-overall-pop)) there is more disagreement when they are broken down by 5 year age groups (Figure \@ref(fig:plot-age-strat)). For those at working age both data sources are comparable (with approximately a 1% difference across all years), however for children, young adults, and those who are 85+ the LFS underestimates the total population. This is particularly the case for older adults with between a 5% and 20% discrepancy for those aged 85-89 and a 25% to 45% discrepancy between those aged 90+. This could be problematic as these age groups often have the most severe outcomes to TB infection. A pragmatic approach to this is to exclude those aged 90+ from future analysis as results for this age group will be subject to large amounts of uncertainty which will be difficult to directly incorporate into the results.

```{r plot-overall-pop, fig.cap = "Overall population estimates derived using Office for National Statistics (ONS) and LFS demographic data. The ONS data is likely to be more reliable as the LFS data is derived using a weighted survey. After accounting for missing UK birth status both datasets provide comparable estimates of the population of England, with a clearly increasing trend over time. However the ONS data indicates a reduction in population from 2000 until 2001 that is not seen in the LFS data. The UK born population has also risen slowly from 2000 until 2015, although the biggest increase has been in the non-UK born population."}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-overall-pop.png")
```

```{r plot-age-strat, fig.cap = "Percentage difference between Office for National Statistics (ONS) population estimates and estimates derived from the LFS by 5 year age group. For most age groups there is less than a 2\\% difference over time. In older adults (85+) there is a substantially greater difference ranging from 5\\% to 40\\%."}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-age-strat.png")
```

## TB notifications

@@

```{r plot-overall-nots, fig.cap = "Notifications in England from 2000 to 2015 stratified by UK birth status, sourced from the ETS system. Notifications in the non-UK born doubled from 3329 in 2000 to 6021 in 2011 since when they have decreased year on year. In the UK born notifications have remained comparable over time, with some evidence of a decrease from 2011 until 2015. UK birth status has become increasingly complete over time with notifications without birth status dropping from 885 in 2000 to 121 in 2015."}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-overall-nots.png")
```

@@

```{r age-dist-cases-nots, fig.cap = "Proportion of cases by 5 year age group in the ETS system in 2005, 2010 and 2015 stratified by UK birth status. Non-UK born cases have a higher proportion of young adult cases with very few cases in children or in older adults. UK born cases have a more uniform distribution of cases with some evidence of a higher proportion of cases in young adults. In the non-UK born the proportion of cases in young adults has decreased over time, with no evidence of a temporal trend in the UK born. These results are not adjusted for population demographics and therefore may be biased."}
if (regen_results) {

@@

## TB incidence rates {#tb-inc-rates-epi}

@@

In the the ETS system, I found a high degree of missing data for several important variables. I also found that there is likely to be strong missing at random (MAR) mechanism underlying this missing data for multiple variables. Several factors are strongly associated with data being missing for many variables, including UK birth status, ethnic group, socio-economic status and year. These MAR mechanisms must be adjusted for in future analysis to avoid bias. I found that date variables in particular suffered from changing data completeness over time, which may introduce spurious temporal trends if not fully understood. I also found that for several variables, including the date of symptom onset, that there was a large degree of recall bias both on the scales of days and months. Several variables, including date of notification and date of starting treatment showed a seasonal trend, with a maximum in the summer months. The date of ending treatment showed less evidence of a seasonal trend but there was some evidence of a maximum number of cases completing treatment in the winter months.

@@

- TB incidence rates stratified by age and UK birth status have been calculated, along with case rates for TB outcomes. These estimates were then to extensively explore trends in TB in England, identifying possible analysis questions to be addressed later in this thesis.

### 2019-07-04 rmds/05-evidence-policy-change.Rmd
checked accroynm usage

Based on these estimated notification rates Sutherland et al. then sought to quantify the ongoing risk of developing notified tuberculosis, projected forward in time, for both the vaccinated and unvaccinated populations. To construct these estimates several key assumptions, based on the results seen in the previous surveys, were made. Firstly, it was assumed that efficacy was not degrading within the ethnic White population and therefore historic estimates would continue to apply into the future. Additionally, it was assumed that the decay of 9% in notification rates, across all ethnic White populations, would continue indefinitely.

@@

I considered the original assumption of a 9% annual decrease in incidence rates as well as three scenarios based on those considered by the JVCI BCG subgroup:[@JVCIBCG2002; @JVCIBCG2003] these were a 3.9% decrease, a 1.9% decrease, and no change annual in incidence rates. I also estimated the annual decrease in incidence rates in the ethnic White UK born using two proxy measures. The first proxy measure was the annual change in notifications in England and Wales, which was estimated using data from Public Health England (PHE). The standard deviation (SD) of this measure was then calculated using the `prop.test` function in R.[@R] The second proxy used was the annual decrease in the UK born age-specific incidence rates in the English population. These were calculated using notification data from the Enhanced Tuberculosis surveillance system (ETS) and the June Labour Force Survey (see Chapter \@ref(data)).[@PHE2017] Incidence rates (with SDs) were estimated using the epiR package.[@EpiR]. Uncertainty was incorporated by sampling from a normal distribution for both proxy measures. Data collection for the ETS began in 2000, I therefore estimated incidence rates between 1984 and 1999, and for the years between notifications surveys (1974-1977 and 1979-1892), using local regression for each sample fitted to the incidence rates published in [@Sutherland1989], and the estimated incidence rates from 2000 on-wards. For years prior to 1973 the annual decreases were assumed to be the mean of the previous 3 years of data. For both proxy measures the annual decreases in incidence rates post 2016 were assumed to be the average of the estimates in 2013, 2014, and 2015.

@@

```{r graph-vacs-to-prev, fig.cap = "Vaccines required in a cohort of those vaccinated at school-age to prevent a single case of Tuberculosis within 15 years of vaccination in 2004, 2009, 2014, or 2019. The years presented were dictated by the 5 year timestep of the model. The percentage annual decrease scenarios considered were based on those considered by the JVCI BCG subgroup, with the addition of a scenario using aggregate notification data and a scenario using estimates of age-specific incidence rates in the UK born. Each boxplot summarises the output of 10,000 model simulations for each scenario. Outliers have been omitted for clarity."}
knitr::include_graphics("chapters/evidence-policy-change/graph-vacs-to-prev-1.png")
```

```{r vac-prevent-tab}
  read_formated("tab-vac-prevent") %>% 
  kable(caption = "The median number (with the 2.5\\% and 97.5\\% quantiles) of vaccines required to prevent a single case of Tuberculosis within 15 years in a ethnic White UK born adult vaccinated at 13 years old.",
               booktabs = TRUE) %>% 
  landscape() %>% 
  kable_styling(latex_options = "hold_position") %>% 
  kable_styling(font_size = 8) %>% 
  column_spec(2, width = "2cm")
```

@@

* I found a 1.9% annual decrease in Tuberculosis incidence rates best matched estimates from notifications.

>>>>

Based on these estimated notification rates Sutherland et al. then sought to quantify the ongoing risk of developing notified TB, projected forward in time, for both the vaccinated and unvaccinated populations. To construct these estimates several key assumptions, based on the results seen in the previous surveys, were made. Firstly, it was assumed that efficacy was not degrading within the ethnic White population and therefore historic estimates would continue to apply into the future. Additionally, it was assumed that the decay of 9% in notification rates, across all ethnic White populations, would continue indefinitely.

@@

I considered the original assumption of a 9% annual decrease in incidence rates as well as three scenarios based on those considered by the JVCI BCG subgroup:[@JVCIBCG2002; @JVCIBCG2003] these were a 3.9% decrease, a 1.9% decrease, and no change annual in incidence rates. I also estimated the annual decrease in incidence rates in the ethnic White UK born using two proxy measures. The first proxy measure was the annual change in notifications in England and Wales, which was estimated using data from Public Health England (PHE). The standard deviation (SD) of this measure was then calculated using the `prop.test` function in R.[@R] The second proxy used was the annual decrease in the UK born age-specific incidence rates in the English population. These were calculated using notification data from the Enhanced TB surveillance system (ETS) and the June Labour Force Survey (see Chapter \@ref(data)).[@PHE2017] Incidence rates (with SDs) were estimated using the epiR package.[@EpiR]. Uncertainty was incorporated by sampling from a normal distribution for both proxy measures. Data collection for the ETS began in 2000, I therefore estimated incidence rates between 1984 and 1999, and for the years between notifications surveys (1974-1977 and 1979-1892), using local regression for each sample fitted to the incidence rates published in [@Sutherland1989], and the estimated incidence rates from 2000 on-wards. For years prior to 1973 the annual decreases were assumed to be the mean of the previous 3 years of data. For both proxy measures the annual decreases in incidence rates post 2016 were assumed to be the average of the estimates in 2013, 2014, and 2015.

@@

```{r graph-vacs-to-prev, fig.cap = "Vaccines required in a cohort of those vaccinated at school-age to prevent a single case of TB within 15 years of vaccination in 2004, 2009, 2014, or 2019. The years presented were dictated by the 5 year timestep of the model. The percentage annual decrease scenarios considered were based on those considered by the JVCI BCG subgroup, with the addition of a scenario using aggregate notification data and a scenario using estimates of age-specific incidence rates in the UK born. Each boxplot summarises the output of 10,000 model simulations for each scenario. Outliers have been omitted for clarity."}
knitr::include_graphics("chapters/evidence-policy-change/graph-vacs-to-prev-1.png")
```

```{r vac-prevent-tab}
  read_formated("tab-vac-prevent") %>% 
  kable(caption = "The median number (with the 2.5\\% and 97.5\\% quantiles) of vaccines required to prevent a single case of TB within 15 years in a ethnic White UK born adult vaccinated at 13 years old.",
               booktabs = TRUE) %>% 
  landscape() %>% 
  kable_styling(latex_options = "hold_position") %>% 
  kable_styling(font_size = 8) %>% 
  column_spec(2, width = "2cm")
```

@@

* I found a 1.9% annual decrease in TB incidence rates best matched estimates from notifications.
